---
title: "Recitation for Week10"
author: "Apoorva Joshi"
format:
  revealjs:
    embed-resources: true
    self-contained-math: true
editor: visual
execute: 
  echo: true
editor_options: 
  chunk_output_type: console
---

## Introduction

```{=html}
<script type="text/x-mathjax-config">
MathJax.Hub.Register.StartupHook("TeX Jax Ready",function () {
  MathJax.Hub.Insert(MathJax.InputJax.TeX.Definitions.macros,{
    cancel: ["Extension","cancel"],
    bcancel: ["Extension","cancel"],
    xcancel: ["Extension","cancel"],
    cancelto: ["Extension","cancel"]
  });
});
</script>
```
-   Supervised Learning is geared toward choosing a modeling procedure that predicts future (proxied by testing) data better than MLE by adding a penalty to the (-) log-likelihood
-   Bayesians can also generate predictions of future data that are even better calibrated than Supervised Learning:
    -   Bayesian propagate uncertainty in $\boldsymbol{\theta}$ through to $Y \mid \boldsymbol{\theta}$

    -   Bayesians can use leave-one-out cross-validation (equivalent to $K$-fold with $K = N$ but can estimate the ELPD without refitting $N$ times) to choose among models

## Okun's Law Revisited

```{r}
source(file.path("..", "Week07", "macroeconomic_data.R")) # do this!
training <- filter(data, quarter_startdate >= "1970-01-01",
                   quarter_startdate <= "1994-10-01")  # 99 quarters
testing <- filter(data, quarter_startdate > "1994-10-01",
                  quarter_startdate < "2019-10-01")    # 99 quarters
```

```{r}
#| message: false
library(rstanarm)
options(mc.cores = parallel::detectCores())
```

```{r}
post <- stan_glm(GDO ~ x, data = training, seed = 12345,
                 prior_intercept = normal(3, 0.5), # on mu
                 prior = normal(-2, 1),            # on beta
                 prior_aux = exponential(0.5))     # on sigma
coef(post) # posterior medians of 4000 estimates (not "the" estimate)
apply(as.matrix(post), MARGIN = 2, FUN = mad) # not that small
```

With only $99$ observations in `training`, we cannot be certain about $\boldsymbol{\theta}$ in Okun's Law (`x` is the change in unemployment).

## Expected Log Predictive Density

$$\mbox{ELPD} = \sum_{n = 1}^N \int_\Omega \ln \int_\Theta f\left(y_{N + n} \mid \boldsymbol{\theta}\right) f\left(\boldsymbol{\theta} \mid \mathbf{y}\right) d\boldsymbol{\theta} dy_{N + n}$$

```{r}
log_lik_testing <- log_lik(post, newdata = testing) # 4000 x 99
loo::elpd(log_lik_testing) # estimating the inner integral with draws
```

The estimate is of an expectation over $\boldsymbol{\theta}$ but is evaluated at the realized $N$ observations in the `testing` data. It does not take the expectation over future data that we haven't observed yet.

## Estimated ELPD

We can *estimate* the ELPD with $\sum\limits_{n = 1}^N \ln f\left(y_n \mid \mathbf{y}_{-n}\right)$, under an additional testable assumption (satisfied in this case) that none of the $N$ observations we conditioned on has an outsized influence on the posterior distribution.

```{r}
loo(post) # only utilizes training data, not testing
```

## Differences

Why is $-216.2$ (`elpd`) different from $-235.7$ (`elpd_loo`)?

. . .

They are not actually that different because the (estimated) standard error of the difference between them is almost $11$. `elpd_loo` yields an (estimated) expectation over all ways that $99$ future observations could be drawn from a normal distribution implied by Okun's Law. `elpd(log_lik(post, newdata = testing))` only uses one of those ways that $99$ future observations could be drawn from a normal distribution implied by Okun's Law. Supervised learners see the latter is "genuine" and forget that past testing data is only a proxy for the unobserved future data they actually care about predicting.

## What You Should Have Done

Only using the `training` data in `stan_glm(GDO ~ x, data = training, …)` wasted the $99$ observations in `testing`, which makes our posterior uncertainty about $\boldsymbol{\theta}$ larger and our predictions of future GDO worse

```{r}
post <- update(post, data = data, # not separating training / testing
               subset = quarter_startdate < "2020-01-01") # pre-covid
(loo_post <- loo(post, save_psis = T)) # for 199 future observations
```

## Model Comparison

```{r}
AR1 <- stan_glm(GDO ~ lag(GDO), data = data, seed = 12345,
                subset = quarter_startdate > "1970-01-01" &
                  quarter_startdate < "2020-01-01",
                 prior_intercept = normal(3, 0.5), # on mu
                 prior = normal(0.5, 0.2),         # on beta
                 prior_aux = exponential(0.5))     # on sigma
loo_compare(loo(post), loo(AR1))
```

. . .

Only using `lag(GDO)` from the previous quarter is expected to be much worse at predicting future `GDO` than is using only `x`, the change in unemployment during the current quarter (although you could use both). We reached that conclusion without wasting a good chunk of our data by separating `testing` from `training` like supervised learners do.

## Model Checking with LOO Concepts

```{r}
pp_check(post, plotfun = "loo_intervals", # predictions worse in 1970s
         psis_object = loo_post$psis_object) # than in later decades
```

## An Example of Decent Calibration

```{r}
pp_check(post, plotfun = "loo_pit_overlay", # too few tail predictions
         psis_object = loo_post$psis_object) # but could be much worse
```

## Conclusion

-   Supervised learning is not the only way to generate predictions for (future) data; just the most popular
-   Supervised learning usually yields better point predictions than Frequentist point predictions (because of penalization) but does not yield a posterior predictive distribution
-   All the things supervised learners do with (functions of) $\widehat{\boldsymbol{\theta}}$ can be done better with each MCMC draw from $\boldsymbol{\theta} \mid \mathbf{y}$
-   Supervised learning splits into training and testing in order to compare models, but that is actively harmful in a Bayesian analysis so you should not do it
